---
title: "HW4"
output: html_document
---

```{r hw3_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

```{r}
library(tidymodels)
library(dplyr)
library(ggplot2)
library(rpart.plot)
library(tidymodels)
library(cluster)
tidymodels_prefer()
flights <- read.csv("D:\\STAT 253\\HW\\FlightData.csv")
```


```{r}
flights_sub <- flights %>% 
  filter(ARR_DELAY_NEW > 0) %>%
  mutate(DEP_HOUR = round(as.numeric(DEP_TIME)/100)) %>%
  mutate(ARR_HOUR = round(as.numeric(ARR_TIME)/100)) %>%
  mutate(ARR_DELAY_LOG = log(ARR_DELAY_NEW)) %>%
  mutate(WEATHER = (WEATHER_DELAY > 5) & !is.na(WEATHER_DELAY)) %>%
  mutate(CARRIER = (CARRIER_DELAY > 5)&!is.na(CARRIER_DELAY)) %>%
  mutate(NAS = (NAS_DELAY > 5)&!is.na(NAS_DELAY)) %>%
  mutate(SECURITY = (SECURITY_DELAY > 5)&!is.na(SECURITY_DELAY)) %>%
  mutate(LATE_AIRCRAFT = (LATE_AIRCRAFT_DELAY > 5) &!is.na(LATE_AIRCRAFT_DELAY)) %>%
  mutate(arr_delay = ARR_DELAY_NEW > 10) %>% 
  left_join(tibble(state.region,state.abb,state.division),by=c('ORIGIN_STATE_ABR' = 'state.abb')) %>%
  rename('ORIGIN_REGION' = state.region) %>%
  rename('ORIGIN_DIVISION' = state.division ) %>%  
  left_join(tibble(state.region,state.abb,state.division),by=c('DEST_STATE_ABR' = 'state.abb')) %>%
  rename('DEST_REGION' = state.region  ) %>%
  rename('DEST_DIVISION' = state.division ) %>% 
  mutate(arr_delay = factor(arr_delay),
         YEAR = factor(YEAR),
         MONTH = factor(MONTH),
         DAY_OF_MONTH = factor(DAY_OF_MONTH),
         DAY_OF_WEEK = factor(DAY_OF_WEEK),
         ORIGIN = factor(ORIGIN),
         ORIGIN_STATE_ABR = factor(ORIGIN_STATE_ABR),
         DEST = factor(DEST),
         DEST_STATE_ABR = factor(DEST_STATE_ABR),
         WEATHER = factor(WEATHER),
         CARRIER = factor(CARRIER),
         NAS = factor(NAS),
         SECURITY = factor(SECURITY),
         LATE_AIRCRAFT = factor(LATE_AIRCRAFT),
         ORIGIN_REGION = factor(ORIGIN_REGION),
         ORIGIN_DIVISION = factor(ORIGIN_DIVISION),
         DEST_REGION = factor(DEST_REGION),
         DEST_DIVISION = factor(DEST_DIVISION),
         OP_CARRIER = factor(OP_CARRIER),
         ORIGIN_STATE_NM = factor(ORIGIN_STATE_NM),
         DEST_STATE_NM = factor(DEST_STATE_NM)) %>%
  select(-X,-(CARRIER_DELAY:LATE_AIRCRAFT_DELAY), -ARR_DEL15) %>%
  na.omit()
```


```{r}
set.seed(123)
flights_sub <- sample_n(flights_sub, 1000)
```

Reseach question: Where is the destination region for certain aircraft?

# 01 - Decision Tree
```{r}
## Model Spec

ct_spec <- decision_tree() %>%
  set_engine(engine = 'rpart') %>%
  set_args(cost_complexity = 0.008,
           min_n = 5, 
           tree_depth = NULL) %>% 
  set_mode('classification')

flight_rec <- recipe(arr_delay ~ DEP_HOUR+ARR_HOUR+WEATHER+SECURITY+DAY_OF_WEEK+NAS+LATE_AIRCRAFT+ ORIGIN_REGION+OP_CARRIER, data = flights_sub) %>%
  #step_unknown(all_nominal_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

flight_wf <- workflow() %>%
  add_model(ct_spec) %>%
  add_recipe(flight_rec)
```


```{r}
## Fit the model
fit_flight <- flight_wf %>%
  fit(data = flights_sub)
```


```{r}
fit_flight %>%
  extract_fit_engine() %>%
  rpart.plot()
```


```{r}
set.seed(123)
flight_fold <- vfold_cv(flights_sub, v=10)

flight_wf_tune <- flight_wf %>%
  update_model(ct_spec %>% set_args(cost_complexity = tune(), min_n = 5000))

param_grid <- grid_regular(cost_complexity(range = c(-4, 1)), levels = 10) # try different values for levels

tune_res <- tune_grid(
  flight_wf_tune, 
  resamples = flight_fold, 
  grid = param_grid, 
  metrics = metric_set(accuracy, sens, yardstick::spec, roc_auc)
)

autoplot(tune_res) + theme_classic()

best_complexity <- select_best(tune_res, metric = 'accuracy')
flight_wf_final <- finalize_workflow(flight_wf_tune, best_complexity)

flight_final_fit <- fit(flight_wf_final, data = flights_sub) # fit final tuned model to training data

flight_final_fit %>% extract_fit_engine() %>% rpart.plot()

# CV Metrics for Tuned Classification Tree
tune_res %>% 
  collect_metrics() %>%
  filter(cost_complexity == best_complexity %>% pull(cost_complexity))
```






# 02 - Random Forest

```{r}
rf_spec <- rand_forest() %>%
  set_engine(engine = 'ranger') %>% 
  set_args(mtry = NULL, # size of random subset of variables; default is floor(sqrt(ncol(x)))
           trees = 50, # Number of bags
           min_n = 5000,
           probability = FALSE, # want hard predictions first
           importance = 'impurity') %>% 
  set_mode('classification') # change this for regression tree

rf_spec


flight_rf_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(flight_rec)
```


```{r}
set.seed(123)
flight_rf_fit <- flight_rf_wf %>%
  fit(data = flights_sub)

flight_rf_fit
```


```{r}
flight_rf_OOB_output <- tibble(
  .pred_class = flight_rf_fit %>% extract_fit_engine() %>% pluck('predictions'),
  DEST_REGION = flights_sub %>% pull(DEST_REGION))

bag_metrics <- metric_set(sens, yardstick::spec, accuracy)

flight_rf_OOB_output %>% 
  bag_metrics(truth = DEST_REGION, estimate = .pred_class)
```


# Soft Prediction
```{r}
set.seed(123) #to get the same bootstrap samples, use same seed
flight_rf_fit2 <- flight_rf_wf %>%
  update_model(rf_spec %>% set_args(probability = TRUE)) %>%
  fit(data = flights_sub)

flight_rf_fit2
```


# Variable Importance
```{r}
library(vip) #install.packages('vip')

flight_rf_fit %>% extract_fit_engine() %>% vip() #based on impurity

flight_rf_wf %>% #based on permutation
  update_model(rf_spec %>% set_args(importance = "permutation")) %>%
  fit(data = flights_sub) %>% extract_fit_engine() %>% vip()
```


Information gained:
We use a decision tree and a random forest to answer where is the destination of certain aircraft. From both methods, we can tell that ORIGIN_REGION_West has the most important predicting power as it is used to be the first split in the tree. The comparison between using a random forest and a decision tree is that, random forest raised the accuracy by 2%. We are thinking adding a logistic regression to answer our classification question despite using trees.




# 03 K-means
```{r}
cust_cluster_sil <- function(k){
    # Perform clustering
    kclust <- pam(daisy(flights_sub), k = k)

    # Return the average silhouette
    return(kclust$silinfo$avg.width)
}

tibble(
    k = 2:15,
    tot_wc_ss = purrr::map_dbl(2:15, cust_cluster_sil)
) %>% 
    ggplot(aes(x = k, y = tot_wc_ss)) +
    geom_point() + 
    labs(x = "Number of clusters",y = 'Average Silhouette') + 
    theme_classic()
```

